Testing Traversability of Protein Tunnels with Flexible Ligands Using Motion Planning

### 2017-06-28 paper.tex
tunnel properties + other comments

\def\C{\mathcal{C}}
\def\CF{\mathcal{C}_{free}}
\def\CFD{{\mathcal{C}^s_{free}}}
\def\dt{d_{tunnel}}
\def\da{d_{atom}}
\def\R{\mathbb{R}}

@@

\def\RRTTD{RRT$_{\mathrm{td}}$}
\def\RRTN{RRT$_{\mathrm{n}}$}
\def\ths{d_\mathrm{surf}}

@@

%\section{Preliminaries}
\section{Traversability of Tunnels}

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
%Let $\SS \subset \mathbf{R}^3$ denote the union of all spheres, i.e., the geometry of the protein.
%We further assume that a subset of atoms located at the surface, are known. These atoms, reffered to as {\sl surface atoms} in this paper,
%   can be detected as atoms located at $\alpha$-shape of the molecule.
%The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
The flexibility of a ligand is modeled using the set $\L$ of its conformations.
The conformations can be prepared, e.g., considering the potential energy.
We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.

@@

\subsection{Main Idea}

The analysis of tunnel traversability is based on computation of multiple trajectories of the ligand inside the tunnel.
As the traversability may depend on the initial position of the ligand, especially its rotation, it is necessary to compute
the trajectories from multiple starting configurations $\qinit$.
The starting configuration $\qinit$ is generated by randomly sampling the vicinity of the first sphere $c_1$ of the tunnel.
A sample $q$ is created by generating its translation part randomly around $t_1$ in the distance $R$.
The rotation part of $q$ and the conformation index are set randomly, the scale is set to $\smin$.
If the sample $q$ is collision-free, it can be considered as a new starting configuration $\qinit$.
Then, multiple trajectories are computed starting from $\qinit$ using the method proposed in the following section.

\subsection{RRT for Ligand Trajectory Computation}

The trajectories are computed using an extension of the RRT~\cite{lavalleRRT} planner.
The RRT planner is extended by guided sampling, scaling of the ligand atoms, and using a predefined set of ligand conformations.
The main loop of the method is described in Alg.~\ref{alg::main}.
%The spheres representing the tunnel are used to generate the random samples for the proposed RRT-based planner.
%We propose a RRT-based method to compute the trajectories of the ligand.
%As the tunnels are usually computed on proteins without ligands, their are usually very narrow and the classic RRT would
%suffer from the well known narrow passage problem~\cite{hannaWIS}.
%To cope with the narrow passage problem, the random samples are not generating from the whole configuration space.
%Instead, they are generated using the spheres of the tunnel.
In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between the $\qrand$ and the tree is performed using the weighted 6D Euclidean metric considering
both 3D rotation and 3D translation.

@@

The result of the planner is the tree $\T$ of collision-free configurations.
The path between $\qinit$ (root of the tree) and $q'$, where $q'$ is the nearest node in the tree towards $\qgoal$, is found.
The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.
Let $d(P,i)$ denote the 3D Euclidean distance.

The path $P$ reaches a particular sphere $t_i$ in the tunnel if the 3D Euclidean distance between the sphere and its nearest configuration
in the path is less than the threshold $T$.
TODO VV dodelam

@@

The traversability of the tunnel $T$ is evaluated using all computed trajectories $P_i$. 
Three basic characteristics of the tunnel are computed from the trajectories: the accessiblity, throughput and the scaling factor.
These characteristics are computed for each sphere $(c_i,r_i) \in T$ of the tunnel.
This requires to determine which parts of the tunnel were accessed by the trajectories.

The Accessibility 
--- The probability of the ligand that it reaches a tunnel's part.
We employ the tunnel representation from CAVER 3.0 and we define a tunnel part as a sphere $P = (x, r)$ where $x \in \mathbb{R}^3$ is sphere center and $r \in \R$ is its radius.
Then, for each trajectory $t$ we define its ending tunnel part $P_t^e$ as the part whose center is closest to the end position of $t$.
Finally, the accessibility of $P$ is defined as a ratio of trajectories that passed $P$ or ended in $P$ over all trajectories.

As described above, our approach enables to estimate the probability that the ligand passes through the tunnel.
On the other hand, the probability does not show any details about the ligand's passage, e.g., how complicated it is to pass through some part of the tunnel, which trajectory is the most used one, etc.
Moreover, it is not feasible to directly visualize and explore the resulting trajectories due to the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
Therefore, we propose to extract and visualize the complex information about the ligand's passage through a tunnel.
In order to further support the tunnel exploration process performed by biochemists.
To do that, we employ an ensemble of resulting trajectories $T$, and we compute and visualize the following tunnel properties:
\begin{itemize}

\item \textbf{Ligand scale} --- The scale of the ligand when it passes a given tunnel's part.
For the trajectory $t$, the ligand scale $s(P_t)$ in the tunnel part $P_t$ is defined as the scale of such sample of $t$ whose position is the closest one to the center of $P_t$.
Then, the average ligand scale is defined as $s(P) = \frac{1}{|T|} \sum_{t \in T} s(P_t)$.
Similarly, the minimum and maximum ligand scales can be defined.
  \item \textbf{Throughput} --- The probability that the ligand passes through a tunnel's part.
The throughput of the tunnel part $P$ is defined as a ratio of trajectories that passed $P$ over all trajectories that reached $P$.
\end{itemize}
In the visualization, these properties are mapped onto the tunnel's surface (see Fig.~\ref{fig:properties}) so that they can be visually correlated with, e.g., the tunnel's shape.
More precisely, a point $x$ on the tunnel surface is colored according to the property value in its nearest tunnel part $P$, i.e., such part $P$ whose center is the closest to $x$ among all tunnel parts.

\begin{figure}
\centering
\begin{tabular}{ccc}
\includegraphics[width=0.3\textwidth]{fig/accessibility}
\quad & \quad
\includegraphics[width=0.3\textwidth]{fig/ligand-scale}
\quad & \quad
\includegraphics[width=0.3\textwidth]{fig/throughput}
\end{tabular}
\caption{Tunnel properties mapped onto its surface.
Left: Accessibility --- green denotes parts that were accessed by the ligand easily while red denotes parts that were accessible only in \tylde 10\% of computations.
Middle: Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
Right: Throughput --- green denotes parts that are highly probable to be passed by the ligand while red denotes those that were hard to pass.}
\label{fig:properties}
\end{figure}

@@

Moreover, we analyze the whole set of trajectories in order to convey the most important trajectories that enabled the ligand to pass or that prevented it from passing.
In order to visualize the important trajectories we first cluster the trajectories according to their similarity.
We define the similarity between two trajectories as follows.
First, we represent each trajectory $t$ by a vector $v = (x_1, x_2, ..., x_N)$ of $N$ samples where $x_i \in \mathbb{R}^3$.
In order to obtain the samples, we find the average starting point $x_s$ of all trajectories and the most distant (w.r.t. $x_s$) ending point $x_e$ also among all trajectories.
Using $x_s$ and $x_e$ we define spheres $S_1, ..., S_N$ such that $S_i = (x_s, r_i)$ where $r_i = i \frac{|x_s x_e|}{N + 1}$.
Then, a vector $v$ for a trajectory $t$ is obtained such that an $i$-th sample $x_i$ in $v$ corresponds to the position in $t$ which is the nearest one to the surface of $S_i$.
Finally, we define a metric $d$ for trajectories $t_i$ and $t_j$ such as $d(t_i, t_j) = \frac{1}{N} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
The clustering is then computed using the UPGMA clustering algorithm~\cite{sokal1958statistical}.
When visualizing, we choose a representative from each cluster and depict its trajectory in 3D using a polyline.
Moreover, we map the representative's cluster size onto the width of the polyline.
In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.4\textwidth]{fig/trajectories-all}
\qquad & \qquad
\includegraphics[width=0.4\textwidth]{fig/trajectories-clustered-21} \\
\end{tabular}
\caption{Visualization of resulting trajectories.
Left: All trajectories colored according to whether they reached the end of the tunnel (green) or not (red).
Right: Trajectories clustered according to their proximity colored by distinct colors.
The different line widths represent the number of trajectories in a cluster.}
\label{fig:trajectories}
\end{figure}

>>>>

\def\C{\mathcal{C}}
\def\CF{\mathcal{C}_{free}}
\def\CFD{{\mathcal{C}^s_{free}}}
\def\dt{d_{tunnel}}
\def\da{d_{atom}}
\def\R{\mathbb{R}}
\def\QI{Q_{init}}

@@

@@

\section{Traversability of Tunnels}

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
The flexibility of a ligand is modeled using the set $\L$ of its conformations.
The conformations can be prepared, e.g., considering the potential energy.
We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.

@@

\subsection{Computing initial configuration}

The analysis of tunnel traversability is based on computation of multiple trajectories of the ligand inside the tunnel, which
requires to generate a set $\QI$ of collision-free initial configurations.
A sample $q$ is created by generating its translation part randomly around $c_1$ in the distance $R$.
The rotation part of $q$ and the conformation index are set randomly, the scale is set to $\smin$.
If the sample $q$ is collision-free, it can be considered as a new starting configuration, so $q$ is added to $\QI$.
For each starting configuration $\qinit \in \QI$, we compute multiple trajectories.
Each trajectory is computed using a variant of RRT, which is proposed in the following section.

\subsection{Computing single trajectory of the ligand}

The trajectories are computed using an extension of the RRT~\cite{lavalleRRT} planner.
The RRT planner is extended by guided sampling, scaling of the ligand atoms, and using a predefined set of ligand conformations.
The main loop of the method is described in Alg.~\ref{alg::main}.
In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between the $\qrand$ and the tree is performed using the weighted 6D Euclidean metric considering
both 3D rotation and 3D translation.

@@

The result of each planning trial is the tree $\T$ of collision-free configurations.
The path between $\qinit$ (root of the tree) and $q'$, where $q'$ is the nearest node in the tree towards $\qgoal$, is found in the tree.
The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.

@@

The trajectories are computed for all initial configurations $\qinit \in \QI$ and for each of the starting
configuration, multiple trajecotries are obtained.
For the purpose of this paper, we compute the tunnel characteristics using all computed trajectories.

A trajectory $P$ reaches the tunnel sphere $c_i \in t$ if the 3D Euclidean distances of the 
nearest configuration $q \in P$ towards $c_i$ is less than $r_i$ (radius of the sphere $c_i$).
Let $N(i)$ denote the number of trajectories, that reached $i-$th sphere of the tunnel, $i=1,\ldots,n$.
Three basic characteristics of the tunnel are computed from the trajectories: accessibility, throughput and the scaling factor.

The accessibility $A(i)$ of the sphere $i$ is the probability of reaching the sphere by the computed trajectories,
i.e., $A(i)=N(i) / N$, where $N$ is the total number of all trajectories.
The accessibility  shows how probable is to pass the tunnel upto the sphere $i$.
The  ligand passage may however be strongly affected by the first bottleneck, so the parts of the tunnel located behind
the bottleneck have low accessibility.

To better measure how probable is to pass a tunnel part, we should take into account only trajectories, that even have chance
to visit it, e.g. the trajectories that reached the preceding tunnel part.
The throughput $T(i)$ is the ratio of trajectories that passed sphere $i$ (i.e., visited sphere $i+1$) and reached the sphere $i$,
i.e, $T(i) = N(i+1) / N(i)$. 
The throughput is not computed for the last sphere ($i=N$).
The throughput shows a local accessibility of the tunnel parts.

%The probability of the ligand that it reaches a tunnel's part.
%We employ the tunnel representation from CAVER 3.0 and we define a tunnel part as a sphere $P = (x, r)$ where $x \in \mathbb{R}^3$ is sphere center and $r \in \R$ is its radius.
%Then, for each trajectory $t$ we define its ending tunnel part $P_t^e$ as the part whose center is closest to the end position of $t$.
%Finally, the accessibility of $P$ is defined as a ratio of trajectories that passed $P$ or ended in $P$ over all trajectories.
%On the other hand, the probability does not show any details about the ligand's passage, e.g., how complicated it is to pass through some part of the tunnel, which trajectory is the most used one, etc.
%Moreover, it is not feasible to directly visualize and explore the resulting trajectories due to the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.

The proposed planner is allowed to scale down radii of the ligand up to the scale $\smin$. 
It can be expected, that narrow parts of the tunnels are more often passed with a more scaled down ligand, than
the wider parts.
Ligand scale $L(i)$ at the tunnel sphere $i$ is the average scale of ligand that reaches the sphere $i$.
The ligand scale is computed as the average value of scales of nearest configurations of the trajectories towards tunnel sphere
$i$.

%--- The scale of the ligand when it passes a given tunnel's part.
%For the trajectory $t$, the ligand scale $s(P_t)$ in the tunnel part $P_t$ is defined as the scale of such sample of $t$ whose position is the closest one to the center of $P_t$.
%Then, the average ligand scale is defined as $s(P) = \frac{1}{|T|} \sum_{t \in T} s(P_t)$.
%Similarly, the minimum and maximum ligand scales can be defined.

@@

%In order to fursupport the tunnel exploration process performed by biochemists.
The computed properties of the tunnel are presented visually by mapping the accessibility, throughput and scale to the sphere representation
of the tunnel (Fig.~\ref{fig:properties}) using a color map.
Alternatively, the surface representation can be used to visualize the tunnel.
In this case, a 3D point on the tunnel surface is colored according to the property value in its nearest tunnel part $i$,  i.e., such part $i$ whose center $t_i$ is the closest to the point among all tunnel parts.

\begin{figure}
\centering
\begin{tabular}{ccc}
\includegraphics[width=0.3\textwidth]{fig/accessibility} &
\includegraphics[width=0.3\textwidth]{fig/throughput} &
\includegraphics[width=0.3\textwidth]{fig/ligand-scale} \\
  (a) & (b) & (c) \\                     
\end{tabular}
\caption{Tunnel properties mapped onto its surface.
(a) Accessibility --- green denotes parts that were accessed by the ligand easily while red denotes parts that were accessible only in \tylde 10\% of computations.
(b) Throughput --- green denotes parts that are highly probable to be passed by the ligand while red denotes those that were hard to pass.
(c) Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
\label{fig:properties}
}
\end{figure}

The mapping using colors helps to identify difficult part of the tunnels, but it is also necessary to show the raw trajectories.
This is necessary for example if the trajectories detour from a part of the tunnel.
The na\"{i}ve visualization of all trajectories is possible, but too slow if the number of trajectories is in the order of thousands.

In order to visualize the most typical trajectories, we cluster the trajectories according to their similarity.
The number of configurations in the trajectories usually differ due to the stochastic nature of RRT and it also depends on the
initial configurations.

To compare two trajectories (which are sequence of configuration in the configuration space), they are first converted
to a vector $v=(x_1,x_2,\ldots,x_M)$, where $x_i \in \R^3$ is a 3D
point representing the trajectory and $M$ is the parameter determining how precisely we want to represent the trajectories.
To obtain the vector $v$ for a given trajectory $P$, first a set of spheres 
 $S_1, ..., S_N$, where $S_i = (x_s, r_i)$ where $r_i = i \frac{|x_s x_e|}{N + 1}$.
%First, we represent each trajectory $t$ by a vector $v = (x_1, x_2, ..., x_N)$ of $N$ samples where $x_i \in \mathbb{R}^3$.
%In order to obtain the samples, we find the average starting point $x_s$ of all trajectories and the most distant (w.r.t. $x_s$) ending point $x_e$ also among all trajectories.
%Using $x_s$ and $x_e$ we define spheres $S_1, ..., S_N$ such that $S_i = (x_s, r_i)$ where $r_i = i \frac{|x_s x_e|}{N + 1}$.
Then, a vector $v$ for a trajectory $P$ is obtained such that an $i$-th sample $x_i$ in $v$ corresponds to the position in $P$ which is the nearest one to the surface of $S_i$.
Finally, we define a metric $d$ for trajectories $P_i$ and $P_j$ as $d(P_i, P_j) = \frac{1}{N} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
The clustering is then computed using the UPGMA clustering algorithm~\cite{sokal1958statistical}.
When visualizing, we choose a representative from each cluster and depict its trajectory in 3D using a polyline.
Moreover, we map the representative's cluster size onto the width of the polyline.
In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.4\textwidth]{fig/trajectories-all}
\includegraphics[width=0.4\textwidth]{fig/trajectories-clustered-21} \\
(a) & (b) \\                       
\end{tabular}
\caption{Visualization of resulting trajectories.
(a) All trajectories colored according to whether they reached the end of the tunnel (green) or not (red).
(b) Trajectories clustered according to their proximity colored by distinct colors.
The different line widths represent the number of trajectories in a cluster.
\label{fig:trajectories}
}
\end{figure}
